---
title: Assortative mating
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(simulateGP)
	library(MASS)
}))
```

What level of bias do we expect when performing MR on a population that has assortative mating, when the assortment correlation is known?



Create parents without matching first, and just one SNP for now




```{r}
generate_individuals <- function(n, maf = 0.5, bgx = 0.5, bgy = 0.5, bxy = 0.3, bux=0.1, buy=0.1)
{
	nm <- n
	nf <- n
	gxm <- make_geno(nid=nm, nsnp=1, af=maf)
	gym <- make_geno(nid=nm, nsnp=1, af=maf)
	um <- rnorm(n)
	xm <- make_phen(eff=c(bgx, bux), indep=cbind(gxm, um))
	ym <- make_phen(eff=c(bxy, bgy, buy), indep=cbind(xm, gym, um))

	gxf <- make_geno(nid=nf, nsnp=1, af=maf)
	gyf <- make_geno(nid=nf, nsnp=1, af=maf)
	uf <- rnorm(n)
	xf <- make_phen(eff=c(bgx, bux), indep=cbind(gxf, uf))
	yf <- make_phen(eff=c(bxy, bgy, buy), indep=cbind(xf, gyf, uf))

	return(tibble(gxm, gxf, gym, gyf, xm, xf, ym, yf, um, uf))
}
```

Match males and females to make assorted spouse pairs. Using a matching function that generates an ordering of males and females that induces a pre-specified correlation, `rho` between the phenotypes that are being assorted on:

```{r}
generate_rho <- function(m, f, rho)
{
	stopifnot(length(m) == length(f))
	require(MASS)
	mvdat <- mvrnorm(n = length(m), mu=c(0,0), Sigma=matrix(c(1,rho,rho,1), 2,2))
	rm <- rank(mvdat[ , 1], ties.method = "first")
	rf <- rank(mvdat[ , 2], ties.method = "first")
	m_order <- order(m)
	f_order <- order(f)
	return(tibble(m = m_order[rm], f=f_order[rf]))
}

match_spouses <- function(dat, rho, matchm, matchf)
{
	matching <- generate_rho(dat[[matchm]], dat[[matchf]], rho)

	dat$gxm <- dat$gxm[matching$m]
	dat$gym <- dat$gym[matching$m]
	dat$xm <- dat$xm[matching$m]
	dat$ym <- dat$ym[matching$m]
	dat$um <- dat$um[matching$m]

	dat$gxf <- dat$gxf[matching$f]
	dat$gyf <- dat$gyf[matching$f]
	dat$xf <- dat$xf[matching$f]
	dat$yf <- dat$yf[matching$f]
	dat$uf <- dat$uf[matching$f]

	return(dat)
}

```

Create children

```{r}
generate_haplotypes <- function(g)
{
	h <- matrix(0,length(g),2)
	h[g == 2,] <- 1
	r <- sample(c(TRUE,FALSE), sum(g == 1), replace=TRUE)
	h[g == 1,][r,1] <- 1
	h[g == 1,][!r,2] <- 1
	stopifnot(all(rowSums(h) == g))
	return(h)
}

create_child_geno <- function(gm, gf)
{
	stopifnot(length(gm) == length(gf))
	tr <- sample(c(TRUE,FALSE), length(gm), replace=TRUE)
	ho <- matrix(0,length(gm), 2)
	hm <- generate_haplotypes(gm)
	hf <- generate_haplotypes(gf)
	ho[tr,1] <- hm[tr,1]
	ho[!tr,1] <- hm[!tr,2]
	ho[tr,2] <- hf[tr,1]
	ho[!tr,2] <- hf[!tr,2]
	return(rowSums(ho))
}

create_child <- function(dat, bgx, bgy, bxy, bux, buy)
{
	dat$gxo <- create_child_geno(dat$gxm, dat$gxf)
	dat$gyo <- create_child_geno(dat$gym, dat$gyf)
	dat$uo <- rnorm(nrow(dat))
	dat$xo <- make_phen(c(bgx, bux), cbind(dat$gxo, dat$uo))
	dat$yo <- make_phen(c(bxy, bgy, buy), cbind(dat$xo, dat$gyo, dat$uo))
	return(dat)
}
```

Run the simulation:

```{r}
bgx <- 0.2
bgy <- 0.8
bxy <- 0.0
bux <- 0
buy <- 0
rho <- 0.8
maf <- 0.2
dat <- generate_individuals(n=1000000, maf=maf, bgx=bgx, bgy=bgy, bxy=bxy, bux=bux, buy=buy)
dat <- match_spouses(dat, rho, "xm", "yf")
dat <- create_child(dat, bgx=bgx, bgy=bgy, bxy=bxy, bux=bux, buy=buy)


# Within male
cor(dat$ym, dat$xm)
cor(dat$xm, dat$gxm)
cor(dat$xm, dat$gym)
cor(dat$ym, dat$gxm)
cor(dat$ym, dat$gym)

# Within female
cor(dat$yf, dat$xf)
cor(dat$xf, dat$gxf)
cor(dat$xf, dat$gyf)
cor(dat$yf, dat$gxf)
cor(dat$yf, dat$gyf)


# Cross spousal phen
cor(dat$xm, dat$yf)
cor(dat$xm, dat$xf)
cor(dat$ym, dat$xf)
cor(dat$ym, dat$yf)


# Cross spousal geno
cor(dat$gxm, dat$gxf)
cor(dat$gxm, dat$gyf)
cor(dat$gym, dat$gxf)
cor(dat$gym, dat$gyf)

# Child
cor(dat$gxo, dat$gyo)
cor(dat$xo, dat$gyo)
cor(dat$gxo, dat$yo)

summary(lm(dat$gxo ~ dat$yo))
summary(lm(dat$gxm ~ dat$gym))
summary(lm(dat$gxm ~ dat$gyf))
summary(lm(dat$gxm ~ dat$xf))
summary(lm(dat$gxm ~ dat$uf))


```


Expected covariance between gxm and gyf

$$
cov(gxm, gyf) = \rho cov(gxm, xm) cov(gyf, yf)
$$

Check:

```{r}
rho * cov(dat$gxm, dat$xm) * cov(dat$gyf, dat$yf)
cov(dat$gxm, dat$gyf)
```

Expected covariance between gxo and gyf


$$
\begin{aligned}
cov(gxo, gyf) &= cov(gxm, gyf) \cdot cor(gxm, gxo) \\
              &= \frac{1}{2} \cdot \rho \cdot cov(gxm, xm) \cdot cov(gyf, yf) \cdot cov(gxm, gxo)
\end{aligned}
$$

Check:

```{r}
cor(dat$gxo, dat$gyf)
cor(dat$gxm, dat$gyf) * cor(dat$gxm, dat$gxo)

cov(dat$gxo, dat$gyf)
cov(dat$gxm, dat$gyf) * cor(dat$gxm, dat$gxo)
cov(dat$gxm, dat$gyf) * 0.5
rho * cov(dat$gxm, dat$xm) * cov(dat$gyf, dat$yf) * 0.5
```


Expected covariance between gxo and gyo

$$
\begin{aligned}
cov(gxo, gyo) &= 2 \cdot cov(gxm, gyf) \cdot cov(gxm, gxo) \cdot cov(gyf, gyo) \\
              &= 4 \cdot \rho \cdot cov(gxm, xm) \cdot cov(gyf, yf) \cdot cov(gxm, gxo) \cdot cov(gyf, gyo) \\
              &= \frac{1}{4} \cdot \rho \cdot cov(gxm, xm) \cdot cov(gyf, yf)
\end{aligned}
$$

Check:

```{r}
cov(dat$gxo, dat$gyo)
cov(dat$gxm, dat$gyf) * cor(dat$gxm, dat$gxo) * cor(dat$gyf, dat$gyo)
cov(dat$gxm, dat$gyf) / 4
with(dat, rho * cov(gxm, xm) * cov(gyf, yf)) / 4
rho * cov(dat$gxm, dat$xm) * cov(dat$gyf, dat$yf) / 4
```

How does this translate to a causal effect estimate?

```{r}
summary(lm(yo ~ gxo, dat))$coefficients[2,1] / summary(lm(xo ~ gxo, dat))$coefficients[2,1]
```


Finally, what is the relationship between gxo and yo

```{r}
cov(dat$gxo, dat$yo)

cov(dat$gxo, dat$gyo) / cor(dat$gyo, dat$yo)^2
rho * cov(dat$gxm, dat$xm) * cov(dat$gyf, dat$yf) / (4 * cor(dat$gyo, dat$yo)^2)

rho * cov(dat$gxm, dat$xm) * cov(dat$gyf, dat$yf) / (4 * cov(dat$gyo, dat$yo)^2 / (var(dat$yo) * var(dat$gyo)))

rho * cov(dat$gxm, dat$xm) * (var(dat$yo) * var(dat$gyo) / (4 * cov(dat$gyo, dat$yo) ))


```


## Other

The child genotype is in HWE

```{r}
chisq.test(
	rbind(
		table(dat$gxo), 
		c(0.25*nrow(dat), 0.5*nrow(dat), 0.25 * nrow(dat))
	)
)
```

Check that the path through the causal effect doesn't influence the assortative mating bias term

```{r}
param <- expand.grid(
	bgx = c(0.1),
	bgy = c(0.6),
	bxy = c(0, 0.1, 0.2),
	bux = c(0.1),
	buy = c(0.1),
	rho = c(0.8),
	sim = 1:10,
	bxyhat = NA
)
rm(bgx,bgy,bxy,bux,buy,rho)

for(i in 1:nrow(param))
{
	message(i, " of ", nrow(param))
	dat <- with(param[i, ], generate_individuals(n=1000000, maf=0.5, bgx=bgx, bgy=bgy, bxy=bxy, bux=bux, buy=buy))
	dat <- with(param[i, ], match_spouses(dat, rho, "xm", "yf"))
	dat <- with(param[i, ], create_child(dat, bgx=bgx, bgy=bgy, bxy=bxy, bux=bux, buy=buy))
	param$bxyhat[i] <- summary(lm(yo ~ gxo, dat))$coefficients[2,1] / summary(lm(xo ~ gxo, dat))$coefficients[2,1]
}

cor(param$bxyhat, param$bxy)
param %>% group_by(bxy) %>% summarise(bxyhat=mean(bxyhat))


```

