---
title: Sib pairs in UKBB
date: 12/11/2018
---


Questions:

1. How many sib pairs are there
2. Is the distribution of sib pairs different from what we expect by chance? (e.g. due to ascertainment)

Full sibs should be IBS of 0.5 (expected) with ~0.36 standard deviation. This means some full sibs look like parent-offspring, and some full sibs (who happen to have little sharing) will look like half-sibs (expected 0.25, with sd of 0.25).

[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3070763/](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3070763/)

Helpfully, the data contains IBS0 (proportion of loci that are unshared between individuals) as well as IBS kinship (average sharing across all loci). Parent offspring will have 0% IBS0, and the IBS0 sharing between half sibs and fullsibs will follow different distributions, so there should be minimal overlap across these two dimensions.


## Analysis

Read in data

```{r}
library(data.table)
a <- fread("../data/ukb8786_rel_s488295.dat")
str(a)
```

```{r}
sum(a$Kinship == -1)
```

Some of the relationships are -1, remove them

```{r}
a <- subset(a, Kinship != -1)
a$Kinship <- a$Kinship * 2 
```

Distribution of all relationships

```{r}
hist(a$Kinship, breaks=200)
hist(a$IBS0, breaks=200)
plot(Kinship ~ IBS0, a)
```

Now we will try to identify the samples within the main central ovoid cluster, which should represent the sibling pairs. 

Try k-means:

```{r}
temp <- cbind(scale(a$Kinship), scale(a$IBS0))

clust <- kmeans(temp, centers=4)
library(ggplot2)
ggplot(a, aes(x=IBS0, y=Kinship)) +
geom_point(aes(col=as.factor(clust$cluster))) +
scale_colour_brewer(type="qual") +
labs(colour="Cluster")

```

This does not work well.

Try gaussian mixture model:

```{r}
# library(ClusterR)
# gmm = GMM(temp, 4, dist_mode = "maha_dist", seed_mode = "random_subset", km_iter = 10, em_iter = 10, verbose = F) 

library(mclust)
temp <- cbind(scale(a$Kinship), scale(a$IBS0))
xyMclust <- Mclust(temp, G=4)
plot(xyMclust)

```

Better but still not good enough.

Try doing it by eye:

```{r}
library(ggplot2)
index <-	a$Kinship > (a$IBS0 * -21 + 0.5) &
			a$Kinship < 0.7 &
			a$IBS0 > 0.001 &
			a$IBS0 < 0.008
ggplot(a, aes(x=IBS0, y=Kinship)) +
geom_point(aes(col=as.factor(index))) +
scale_colour_brewer(type="qual") +
geom_abline(intercept=0.5, slope=-21) +
geom_hline(yintercept=0.7) +
geom_vline(xintercept=0.008) +
geom_vline(xintercept=0.001) +
labs(colour="Full sibs")
```

```{r}
sibs <- subset(a, index)
nrow(sibs)
```

Distribution of sibs

```{r}
hist(sibs$Kinship)
```

Summary of sibs

```{r}
summary(sibs$Kinship)
```

Save the sib pairs

```{r}
a$sibpair <- index
write.csv(a, file="../data/ukbb_sibs.csv")
